---
title: "Does Curriculum Matter?"
author: "Allison Theobold"
date: "September 24, 2019"
output:
  pdf_document: default
  word_document: default
---

# Introduction 

This research investigates the relationship between students' curriculum and 
the grade they received, specifically for Montana State University students 
enrolled in STAT 216. In the spring of 2015 students enrolled in either a 
"traditional" STAT 216 course or a "TEAL" based curriculum. The TEAL curriculum 
was a simulation based, hands-on course, taught in a interactive TEAL 
(Technology Enhanced Active Learning) classroom. This course had daily 
activities from a course pack, and very little lecture. The "traditional" 
curriculum was given in a typical lecture style classroom, with daily lectures,
and select weekly activities similar to the TEAL course.  

Research has shown that these TEAL classrooms increase student engagement and 
retention. Therefore, we wish to study if the TEAL curriculum provides an 
overall better outcome for student's grades in STAT 216. 

## Data Collection 

The data collected on students in STAT 216 in the spring of 2015 include the 
following variables: 

* the student's unique identifier (number)  
* what score they earned on the ACT math exam  
* what score they earned on the SAT math exam    
* what score they earned on the math placement exam (at MSU)   
* when they took their most recent math course (30 = spring, 70 = fall, 50 = summer)   
* the last math course they took (math subject plus math course)  
* the grade they received in that math course  
* similar information for their last statistics course  
* what their AP Calc grade transferred as  
* what their AP Stat grade transferred as  
* what their cumulative GPA was the semester prior to enrolling in Stat 216  
* the curriculum each student was enrolled in (Lock or MSU.0)  
* their Stat 216 course grade, converted to the GPA scale (4.0 = A, 3.0 = B, 2.0 = C, 1.0 = D, etc.)  


# Methods  

For the research question of interest, the explanatory variable refers to the 
curriculum which the students took, either the TEAL or traditional, and the 
response variable refers to the grade which the students received. 


```{r plots2015, echo = FALSE, message = FALSE, fig.width = 5, fig.height = 4}
library(tidyverse)

stat216 <- read.csv("C:/Users/allisontheobold/Documents/Stat 511/Labs/stat-411-511-lab5/data/stat216_spring2015.csv")

ggplot(stat216, mapping = aes(x = curric, y = GPAgrade)) + 
   geom_violin()

```

For the data of Spring 2015 the students in the "traditional" curriculum and the
TEAL curriculum got the same mean grade with distributions looking the same via
a violin plot of the data. We can see this visually on the violin plot  
where the traditional and TEAL curricula have similar shaped distributions with
equal centers and spread. 

The statistical procedures we will be using to determine if there exists a
difference in true mean student GPA between the TEAL curriculum and the 
traditional for the 2015 students is a two-sample t-test on 1192 degrees of 
freedom. To estimate the difference in true means between the two curricula, we
will provide an approximate 95% confidence interval using t-based methods. 

We will proceed with a t-test because we have met the assumptions of the method, 
(1) the data are independent, and (2) the two groups have similar distributions. 


## Hypotheses  

For our research question we will test the following hypothesis, 

$\mu_{traditional} \neq \mu_{TEAL}.$ 


# Results 

After running the statistical analysis we find that there is significant 
evidence to show that the mean grades of students in the TEAL program differ 
from the mean grades of the students in the traditional program 
(p-value = 0, t = -3.9114). there is strong evidence supporting the alternative
hypothesis that the spring 2015 TEAL curriculum has an effect on a student's 
course grade. Due to these results, we conclude that the TEAL curriculum 
increases students' course grade at Montana State University during the spring
semester of 2015.


```{r spring15, message = FALSE}
library(infer)

t.test(GPAgrade ~ curric, data = stat216)

```


# Scope of Inference  

The students in STAT 216 for spring 2015 were not randomly selected from a 
larger population of STAT 216 students, and therefore we cannot infer the 
findings of this research to a larger population. Rather, we can only infer the
results of this study to those who wherein the study. In addition, the students
were not randomly assigned to the different types of curricula, which prohibits
us from being able to draw any cause-and-effect conclusions for this study.


# Code Appendix

```{r, ref.label = "plots2015", eval = FALSE}

```

